A large leading tail of atmospheric escape in the HAT-P-67 b system

### 2023-03-30 paper/paper1/ms.tex
Prose

The observation that \hatpb---the only hot Saturn searched for atmospheric escape to date---exhibits mass loss indicative of a terminal fate leads us to a natural explanation for the dearth of inflated hot Saturns.  The inflationary lifetime of hot Saturns is so short that observing planets of this class is unlikely.  \hatpb~ appears to be one of the rare planets that we happen to catch a glimpse of before it has been completely annihilated.  The lack of inflated hot Saturns is analogous to the small planet radius gap \citep{2017AJ....154..109F}, with Ohmic dissipation and replacing core-powered mass loss as a heating mechanism, and photoevaporation at play.

We therefore attribute this cliff in the $m_p$ vs $R_p$ diagram to Ohmic dissipation.  \citet{2018AJ....155..214T} emphasized the positive feedback loop as a cause for the lack of sub-Saturns, and noted an alternative interpretation-- mass-dependent migration segregation.  While both scenarios are conceivable, the evidence of strong evaporation of \hatpb~ favors the evaporation scenario.  Futhermore, the runaway evaporation of hot-Saturns was naturally predicted by the Ohmic dissipation mechanism, whereas mass-dependent migration has only been developed as a post-diction.

\subsection{Predictions}

As seen in Figure \ref{fig:ThornFortUpdated}, \hatpb has few other known analogs.  The closest analogues appear both smaller and more massive. Nevertheless, we predict that these few other inflated hot-Saturn \hatpb-analogs should also show evidence for atmospheric escape, albeit likely at a lesser scale.  These sources make excellent targets for \ion{He}{1} 10830 observations, and/or other atmospheric escape diagnostics.  The key figure-of-merit can be distilled to $\tau_\mathrm{infl}$, and setting the $\epsilon$ term to unity (Eq. \ref{eqInflate}).  Following \citet{2018PASP..130k4401K} we define a Helium Spectroscopy Metric (HeSM):

\begin{eqnarray}
    \mathrm{HeSM} = \tau_\mathrm{infl} \label{HeSM}
\end{eqnarray}

@@

As emphasized, there are significant uncertainties in the inflation timescale, but its order-of-magnitude value gives us a way to prioritize target selection, since these may be the sources most amenable to detection of atmospheric escape.  Table \ref{tabHeSM} lists $15$ candidates that have nominal lifetimes less than about 40 Myr.

>>>>

The observation that \hatpb---the only hot Saturn searched for atmospheric escape to date---exhibits mass loss indicative of a terminal fate leads us to a natural explanation for the dearth of inflated hot Saturns.  The inflationary lifetime of hot Saturns is so short that observing planets of this class is unlikely.  \hatpb~ appears to be one of the rare planets that we happen to catch a glimpse of before it has been completely annihilated.  The lack of inflated hot Saturns is analogous to the small planet radius gap \citep{2017AJ....154..109F}, with Ohmic dissipation replacing core-powered mass loss as a heating mechanism, and with photoevaporation still at play.

We therefore attribute this cliff in the $M_p$ vs $R_p$ diagram to Ohmic dissipation.  \citet{2018AJ....155..214T} emphasized the positive feedback loop as a cause for the lack of sub-Saturns, and noted an alternative interpretation-- mass-dependent migration segregation.  While both scenarios are conceivable, the evidence of strong evaporation of \hatpb~ favors the evaporation scenario.  Futhermore, the runaway evaporation of hot-Saturns was naturally predicted by the Ohmic dissipation mechanism, whereas mass-dependent migration has only been developed as a post-diction.

\begin{figure}[t]
    \includegraphics[width=\linewidth]{figures/tf2018_fig2_theory.png}
    \caption{Ohmic dissipation runaway inflation timescale in the exoplanet mass-radius diagram.  Individual planets with confident mass and radius detections are shown as small black dots. Helium non-detections are shown in red squares with all detections of atmospheric escape coming from \citet{2022arXiv221116243D}.  \hatpb~ resides in a sparsely populated region with a vanishingly short inflationary timescale, making it unstable to runaway evaporation with large mass loss expected.}
    \label{fig:tf2018_theory}
\end{figure}

As seen in Figure \ref{fig:ThornFortUpdated}, \hatpb~ has few other known analogs.  Figure \ref{fig:tf2018_theory} illustrates why: the inflationary lifetime for planets with comparable mass, radius, and $T_\mathrm{eq}$ is exceptionally short, and so they are rare.  We constructed the $\tau_\mathrm{infl}$ contours in the following way.  We censored the $\tau_\mathrm{infl}$ to regions in the mass-radius plane where numerical simulations of Ohmic dissipation resulted in runaway inflation \citet{2011ApJ...738....1B}.  We interpolate the $T_\mathrm{eq}$ trendlines from Fig. \ref{fig:ThornFortUpdated} to compute $\tau_\mathrm{infl}$ from Equation\ref{eqInflate}, including the best fit Gaussian prescription for $\hat{\epsilon_G}$ from Eq. 34 of \citet{2018AJ....155..214T}.  This trend should exhibit significant uncertainty owing to the order-of-magnitude nature of the scaling relation, source-to-source scatter in $\hat{\epsilon}_G$, and to what extent $\hat{\epsilon}_G$ applies for planets less than $0.5 M_\mathrm{Jup}$, which were excluded in the derivation of $\hat{\epsilon}_G$ owing to their scarcity.  Nevertheless, this contour appears to match the bulk trends: the bulk locus of planets reside in either the non-inflationary, stable scenario, or the inflationary-yet-stable scenario.  A relatively few $\sim$dozens of planets approach the ``cosmic shoreline'' of inflationary lifetimes in the tens of Myr. The detection of atmospheric escape along this frontier (green circles in Figure \ref{fig:tf2018_theory}) support this interpretation.  More observations and theoretical modeling of atmospheric escape along this ridge could help better understand its realized versus expected shape.

\subsection{HAT-P-67b analogues}

The key figure-of-merit can be distilled to $\tau_\mathrm{infl}$, whic can be thought of as an atmospheric escape metric \citep{2018PASP..130k4401K} for inflated planets in the shaded regime of Figure \ref{fig:tf2018_theory}.  Table \ref{tabHeSM} presents the rank-ordered list of planets by this metric, indicating that they are mostly smaller and more massive than \hatpb.  Only KELT-11~b shows a shorter nominal inflationary timescale than \hatpb, owing to its low mass and equilibrium temperature residing closer to the peak of $\epsilon_G(T\mathrm{eq})$.  We predict that these few other inflated \hatpb-analogs should show evidence for significant atmospheric escape, comparable to what we see for \hatpb.  These sources make excellent targets for \ion{He}{1} 10830 observations, and/or other atmospheric escape diagnostics.  As emphasized, there are significant uncertainties in the inflation timescale, but at least its order-of-magnitude value gives us a quantitative and justified way to prioritize target selection.

@@
